Adaptive evolution with aneuploidy and mutation

### 2022-06-01 ms/ms.tex
formatting
added martin's new plots and a basic discussion of these figures

We also checked a model in which aneuploidy occurs but is adaptively neutral compared to the wild type, that is, $w_{\anwt} = w_{\euwt}$ and $w_{\anmt} = w_{\eumt}$ but $\delta>0$. This model fits the data better than the model with no aneuploidy (in which $\delta=0$), but worse than a model with positive selection for aneuploidy, in which $w_{\euwt}<w_{\anwt}<w_{\anmt}<w_{\eumt}$~(\Cref{fig:2n*-fixation}). 
Thus, our results suggest that aneuploidy plays a significant role in the evolutionary dynamics \citep{Yona2012,Yona2015}.

@@

\paragraph*{Genotype frequency dynamics.}
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}). % TODO new figure by Martin and discussion of the probability of 2n* arriving via 2n+1 - transient solution etc.

% Fig dynamics
%% generated with dynamics-plot.ipynb
\begin{figure}[h!]
  \centering
\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
  \caption{
  \textbf{Posterior genotype frequency dynamics for the model.} 
  The posterior prediction for the frequencies of the four genotypes over time. Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.
  }
  \label{fig:ppc-plot}
\end{figure}

@@

\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, and Tal Simon for discussions and comments.
This work was supported in part by 
the Israel Science Foundation (YR 552/19) and
Minerva Stiftung Center for Lab Evolution (YR).
% TODO add Minerva for Martin; funding for Avihu?
}

@@

\section*{Supplementary material}
\beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/

\subsection*{Supplementary analysis}
\label{sec:supp_analysis}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\subsection*{Supplementary figures \& tables}

% Fig Validation of posteriors
%% generated with diff-runs.ipynb
\begin{figure}[p]
    \centering
      \includegraphics[width=0.45\textwidth]{../figures/runs-A.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/runs-B.pdf}    
      \includegraphics[width=0.325\textwidth]{../figures/runs-C.pdf}      
      \includegraphics[width=0.325\textwidth]{../figures/runs-D.pdf}      
      \includegraphics[width=0.325\textwidth]{../figures/runs-E.pdf} 
       \caption{
    \textbf{Posterior distribution validation.}
    The posterior distribution of model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\cref{eq:heatstress-likelihood}).
    }

>>>>

We also checked a model in which aneuploidy occurs but is adaptively neutral compared to the wild type, that is, $w_{\anwt} = w_{\euwt}$ and $w_{\anmt} = w_{\eumt}$ but $\delta>0$. This model fits the data better than the model with no aneuploidy (in which $\delta=0$), but worse than a model with positive selection for aneuploidy, in which $w_{\euwt}<w_{\anwt}<w_{\anmt}<w_{\eumt}$~(\Cref{fig:2n*-fixation}). 
Thus, our results suggest that aneuploidy significantly affects the evolutionary dynamics.

@@

\paragraph*{Competition between mutation and aneuploidy.}

We simulated 50 % TODO is it 50?
 replicate genotype frequency dynamics using the MAP estimate parameters. 
The MAP frequency dynamics (\Cref{fig:ppc-plot}A) show the four genotypes but also the frequencies of \eumt\ arose from either the \anwt\ genotype via mutation and chromosome loss events, or direct from the \euwt\ genotype via mutation.
We observe that \anmt\ never reaches substantial frequency.
But more importantly, we observe that the majority % TODO add number
of \eumt\ cells are descendant of \euwt\ cells and arose via direct mutation, despite the fact that the \anwt\ genotype reaches high frequencies in the population. % TODO add number
This finding is not unique to the MAP parameter estimate. 
We simulated genotype frequency dynamics using 100,000 parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the average frequency of \eumt\ arising from \anwt at the end of the simulation, $F_A$, over 50 replicates. 
$F_A$, was % TODO add number
and only in 681 (0.6\%) of the 100,000 posterior samples was it larger than 0.5
(see \Cref{fig:ppc-plot}B) for the full histogram). 
Overall, the frequency of \eumt\ descending from \anwt\ increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate (\Cref{fig:ppc-plot}C).
The cases in which $F_A>0.5$ can be roughly divided to two clusters.
The first, with a low formation rate ratio ($\delta/\mu \approx 500$) and a small fitness difference between \eumt\ and \anwt\ ($w_{\eumt}/w_{\anwt}-1 \approx 0.0055$). In these cases, aneuploidy is not generated much faster than mutation, but this is compensated by similar fitness values that decrease the fixation probability of \eumt\ and limit the strength of clonal interference.
The second cluster, with high formation rate ratio ($\delta/\mu \approx 10^5$) and a higher fitness difference between \eumt\ and \anwt\ ($w_{\eumt}/w_{\anwt}-1 \approx 0.007$). 
In these cases, the larger fitness difference between \anwt\ and \eumt\ means that once \eumt\ appears it can more easily avoid extinction and compete with \anwt; this is compensated by much faster generation of \anwt\ and can therefore lead to high $F_A$ (note that we only use samples from the posterior, that is, we only compute $F_A$ for parameter choices that reasonably agree with the experimental results.)

%

% Fig dynamics
\begin{figure}[h!]
  \begin{subfigure}{0.5\textwidth}
      \centering
      %% generated with dynamics-plot.ipynb
      \includegraphics[width=\textwidth]{../figures/dynamics.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/histo_100000.pdf}
  \end{subfigure}
  \\
  \begin{subfigure}{0.33\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/scatter100000_log_star.pdf}
  \end{subfigure}
  \begin{subfigure}{0.33\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/scatter_dirmut_log_star_byr.pdf}
  \end{subfigure}
  \begin{subfigure}{0.33\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_fitr.pdf}
  \end{subfigure}
  \caption{
    \textbf{Posterior genotype frequency dynamics.} 
    \textbf{(A)} The posterior prediction for the frequencies of the four genotypes over time. Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.	
     }
  \label{fig:ppc-plot}
\end{figure}

@@

%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, and Tal Simon for discussions and comments.
This work was supported in part by 
the Israel Science Foundation (YR 552/19) and
Minerva Stiftung Center for Lab Evolution (YR).
% TODO add Minerva for Martin; funding for Avihu?
}

@@

\section*{Supplementary Material}
\beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/

\subsection*{Supplementary Analysis}
\label{sec:supp_analysis}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\subsection*{Supplementary Figures \& Tables}

% Fig Validation of posteriors
%% generated with diff-runs.ipynb
\begin{figure}[h]
    \centering
      \includegraphics[width=0.45\textwidth]{../figures/runs-A.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/runs-B.pdf}    
      \includegraphics[width=0.325\textwidth]{../figures/runs-C.pdf}      
      \includegraphics[width=0.325\textwidth]{../figures/runs-D.pdf}      
      \includegraphics[width=0.325\textwidth]{../figures/runs-E.pdf} 
       \caption{
    \textbf{Posterior distribution validation.}
    The posterior distribution of model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\cref{eq:heatstress-likelihood}).
    }
